%Link back
Now that we have defined our operational semantics (in  Chapter \ref{ch4}) and our denotational semantics (in Chapter \ref{ch6}), we can relate them using the Adequacy Theorem: 

\vspace{1cm}
%Statement of thm
\begin{thm}\label{adeq}
If $\vdash e : \nat$ (i.e.\ $e$ is a closed term of type $\nat$), then $\forall n \in \mathbbm{N}. \ \llbracket e \rrbracket = n \Leftrightarrow e \mapsto^* \underline{n}$
\end{thm}

\vspace{0.5cm}

where $\llbracket e \rrbracket$ = $\llbracket \vdash e : \nat \rrbracket$ and $\underline{n}$ represents the numeral $n$, (as opposed to the actual natural number $n$).

\vspace{0.5cm}

%EXPLAIN WHY ONLY AT BASE TYPE.

%Correctness as in before
The theorem has two directions. We have mostly proved the backwards direction because of the Soundness proof we proved in the previous chapter (see Section \ref{sound}). Therefore we have proved half of Adequacy.

Therefore, the right to left direction of adequacy is a corollary of the soundness proof:

\vspace{0.5cm}

\begin{cor}
If $\vdash e : \nat$ and $\llbracket e \rrbracket = n$ then $e \mapsto^* \underline{n} \Rightarrow \llbracket e \rrbracket = n$
\end{cor}


\begin{proof}
We rewrite $e \mapsto^* \underline{n}$ as $e \mapsto e_0 \mapsto \dots \mapsto e_m \mapsto \underline{n}$ , for any $m \geq 0$. Applying soundness to these evaluations gives us $\llbracket e\rrbracket = \dots = \llbracket e_m \rrbracket = \llbracket \underline{n} \rrbracket$. Therefore we have $\llbracket  e \rrbracket = \llbracket \underline{n} \rrbracket$.

Now we need to prove $\forall n \in \mathbb{N}. \llbracket \underline{n} \rrbracket = n$, which we prove by induction on $n$:

If $n = 0$, then $\llbracket \underline{0} \rrbracket = \llbracket z \rrbracket = 0$, by the definition of the denotational semantics for zero.

%If $\underline{n} = zero$, then by the denotational semantics for $ zero$ we have $\llbracket e_n \rrbracket =  0$.

If $n = n + 1$, then $\llbracket \underline{n + 1} \rrbracket = \llbracket s(\underline{n}) \rrbracket$. By the inductive hypothesis we have $\llbracket \underline{n} \rrbracket = n$, so $\llbracket s(\underline{n}) \rrbracket = n + 1$. The definition of $\llbracket s(\underline{n}) \rrbracket$ is $\llbracket  \underline{n} \rrbracket + 1$. Therefore $\llbracket \underline{n + 1} \rrbracket = n + 1$.

%If $\underline{n} = s(v)$, then  in the rule, we have $\llbracket v \rrbracket$, which equals $v$ by the inductive hypothesis. Therefore, we use the case in the successor rule for a number and get $v + 1$.

Therefore $\forall n \in \mathbb{N}. \llbracket \underline{n} \rrbracket  = n$, so $\llbracket e \rrbracket =  \llbracket \underline{n} \rrbracket  = n$.

\end{proof}

To prove the forwards direction, we could naively try to prove it by induction on each expression. However,  although we only consider closed expressions of base type, if it has sub-expressions then they may not be closed or of type $\nat$, so we have no inductive hypothesis to provide in these cases.

Therefore, instead we use the Logical Relations approach that we described in Section \ref{log}.

%Binary Logical Relation definition
\section{Logical Relation}\label{const}
The logical relation we define is a binary logical relation between denotations of PCF expressions and closed expressions of PCF. It is based on a binary logical relation defined in \citep{Streicher06}. 

There are some differences between our logical relation and the one defined in \citep{Streicher06}. In our proof of the fixpoint case of the main lemma, we do not use the observational preorder that they use. Instead we use an Expansion Lemma (see Lemma \ref{exp} below). We also work with chains instead of directed sets.

Also, Streicher's logical relation is defined on a version of PCF in which  constants to replace certain constructors in the language.

For example, in our definition of PCF, we have terms $s(e)$, for some expression $e$. Here $s$ is a constructor, so $s$ itself is not a term, but once we attach an expression to it, it is a term, $s(e)$.

This is the same for zero, $\case$ and $\fix$.

Alternatively, we could have defined these operations as constants, for example, $\succc : \nat \to \nat$ is a function in the language that has a pre defined definition that never changes. We then also define zero $: \nat$, $\case : \nat \to \nat \to \nat$ (just on Natural numbers) and $\fix : (A \to A) \to A$. This is how PCF is defined in \citep{Streicher06}.
 
\subsection{Defintition of the logical relation}
The definition of the logical relation is the following:

\vspace{0.5cm}

\begin{defn}
We create a family of binary  relations $R_A \subseteq \llbracket A \rrbracket \times \{e \ | \ \vdash e : A \}$, by defining relations by induction on types for each type as follows:

\[ d  R_{\nat}  e \Leftrightarrow \forall n \in \mathbbm{N}. \ d = n \Rightarrow e \mapsto^* n\]

\[f  R_{A \to B}  e \Leftrightarrow \forall d \in \llbracket A \rrbracket. \ \forall e' \in \{ e \ | \ \vdash e : A\}. \ d R_A e' \Rightarrow f(d) R_B  e \ e'\]

\end{defn}

The forwards direction of Adequacy is given by the definition of the relation on expressions of type $\nat$.

As we previously described in section \ref{log}, to prove our actual property, we prove a more general Main Lemma on the logical relation first. Our Main Lemma for this relation will be the following:

\vspace{0.5cm}

\begin{lem} \label{main}
If $\Gamma = x_1 : A_1, \dots, x_n : A_n$, $\Gamma \vdash e : A$ and $ \ d_1 R_{A_1} e_1, \ d_2 R_{A_2} e_2, \dots , \  d_n   R_{A_n} e_n$, then

\[ \llbracket \Gamma \vdash e : A \rrbracket (d_1, \dots, d_n)  \ R_A \  [e_1/x_1, \dots, e_n/x_n]e \]
\end{lem}

In the lemma we are given an expression $e$ that is well typed in a typing context $\Gamma$ and some denotations $d_i$ that are related to some expressions $e_i$.

The denotation of that expression applied to substitutions $(d_1, \dots d_n)$ for all the  free variables in $\Gamma$ will be related to the expression $e$ with the expressions that correspond to each denotation (according to our assumption) substituted for the free variables in $e$.

We will prove this by induction on each possible expression $e$.

%substitution function
\subsection{Substitution Function}
In the Main Lemma, we make multiple substitutions in the expression $e$, but we only defined our substitution function (in Chapter \ref{safe}) on single substitutions $[e'/x]e$. Therefore we define substituting multiple variables in the following way, where $\gamma = [e_1/x_1, \dots e_n/x_n]$:

\begin{adjustwidth}{2cm}{}
$[\gamma](zero) = zero$
\end{adjustwidth}

\begin{minipage}{3.5in}
\begin{align*}
[\gamma](x) = 
  \begin{cases} 
           [\gamma](x) & \text{if } x \in dom(\gamma) \\
           x & otherwise 
  \end{cases}
\end{align*} 
\end{minipage}

This says that if $x$ is present in $\gamma$, (there is a substitution for it), replace $x$ with the result of the substitution in $\gamma$. Otherwise there is nothing to replace $x$ with, so we return the variable unaltered. 

\vspace{0.5cm}

\begin{adjustwidth}{2cm}{}

$[\gamma](s(e)) = s([\gamma] (e))$

$[\gamma](\case (e, z \mapsto e_0, s(v) \mapsto e_S)) = \case ([\gamma](e), z \mapsto [\gamma](e_0), s(v) \mapsto [\gamma](e_S))$

$[\gamma](e \ e') = ([\gamma] e) ([\gamma] e')$

$[\gamma] (\lambda x:A. e) = \lambda x:A. [\gamma] e$

$[\gamma] (\fix \ x:A. e) = \fix \ x:A. [\gamma] e$
\end{adjustwidth}

\vspace{0.5cm}

For a non empty $\gamma = e_1/x_1, \dots e_n/x_n$, we have:

\[ [e_1/x_1, \dots , e_n/x_n] e = [e_1/x_1]([e_2/x_2]( \dots [e_n/x_n] e) \]

If we add a variable to the substitution we can define this in the following way:
\[ [t/x][\gamma] e = [\gamma, t/x] e \]

%Lemmas we need for main lemma
\section{Lemmas for Main Lemma}
We are almost ready to prove the Main Lemma, but first we prove some lemmas that we will use in the proof:


\subsection{Bottom Element Lemma}

\vspace{0.25cm}

\begin{lem} \label{bot}
For any type $A$ and $\Gamma \vdash e: A$, $\bot \ R_A \ e$ 
\end{lem}

\begin{proof}
By induction on types. For $\nat$, we want to show $\forall n \in \mathbbm{N}. \ \bot  = n \Rightarrow e \mapsto^* n$. Because $\bot \notin \mathbbm{N}$, this statement is vacuously true.

For $R_{A \to B}$, we have that $\bot$ is the function $\lambda x : A. \bot$, so we want to show $\forall  d \in \llbracket A \rrbracket. \ \forall e' \in \{ e \ | \ \vdash e : A\}. \ d R_A e' \Rightarrow (\lambda x : A. \bot)(d) R_B  e \ e'$. Assume $d$ is an element of the domain representing type $A$ and $e'$ is an expression of type $A$ such that $d \ R_A e'$. We want to show that $\bot \ R_B \ e \ e'$ and by the inductive hypothesis, we know for any $e : B$ that $\bot \ R_B \ e$, so we can just use this with $e \ e'$ to get our conclusion.
\end{proof}

\subsection{Expansion Lemma}\label{exp}

\vspace{0.25cm}

\begin{lem}\label{exp}
If $\Gamma \vdash e : A$ and $e \mapsto e'$ and $d R_A e'$ then $d R_A e$ 
\end{lem}

\begin{proof}
By induction on types. For base case we assume $d R_{\nat} e'$, so we have $\forall n \in \mathbbm{N}. \ d = n \Rightarrow e' \mapsto^* n$. Let $n = d$. Then, as $e \mapsto e'$ and $e' \mapsto^* n$, we know that $e \mapsto^*n$. Therefore, $d R_{\nat} e$.

For the inductive case, assume $e_0 \mapsto e'_0$ and $f \ R_{A \to B} \ e_0$, so we have $\forall a \in \llbracket A \rrbracket. \ \forall e' \in \{ e \ | \ \vdash e : A\}. \ a \ R_A \ e' \Rightarrow f(a) \ R_B \ e_0 \ e'$. Let $a$ be an element in the domain representing $A$ and $e_1$ be an expression of type $A$ such that $a R_A e_1$. Then $f(a) \ R_B \ e_0 \ e_1$.

By the inductive hypothesis we know that for any $e \mapsto e'$ for expressions of type $B$, that they are both related to the same denotation. Therefore we have $f(a) \ R_B \ e'_0 \ e_1$, so we know $\forall a \in \llbracket A \rrbracket. \ \forall e' \in \{ e \ | \ \vdash e : A\}. \ a \ R_A \ e' \Rightarrow f(a) \ R_B \ e'_0 \ e'$, so $f \ R_{A \to B} \ e'_0$. 
\end{proof}

\subsection{Chains Lemma}

This lemma was adapted from a similar lemma on directed sets from \citep{Streicher06}.

\vspace{0.25cm}

\begin{lem}\label{chain}
For an expression $e : A$ and a  chain $x_0 \sqsubseteq x_1 \sqsubseteq \dots$, if $x_n R_A e$, then $\bigsqcup x_n \ R_A \ e$
\end{lem}

\begin{proof}
By induction on types. For base type, $\nat$, we assume we have a chain of elements in $\mathbbm{N}_\bot$ such that $x_n \ R_{\nat} e$. Therefore for any element in the chain we know $\forall n \in \mathbbm{N} . \ x_n = n \Rightarrow e \mapsto^* n$. We have two cases depending on the values of $\bigsqcup x_n$:

\begin{enumerate}
\item{$\bigsqcup x_n = n$. Then we know that $e \mapsto^* \sqcup x_n$, as any $x_n R_{\nat} e$}
\item{$\bigsqcup x_n = \bot$. By Lemma \ref{bot} we know that $\bot R_{\nat} e$ for any $e$}
\end{enumerate}

The inductive case is for $R_{A \to B}$. Assume we have a chain $f_1 \sqsubseteq f_2 \sqsubseteq \dots$ of elements in $\llbracket A \rrbracket \to \llbracket B \rrbracket$ and $d \ R_A \ e'$ for some $d \in \llbracket A \rrbracket$ and $e : A$. Then we need to show $\bigsqcup f_n (d) \ R_B \ e \ e'$. By induction we know for any expression of type $B$ and a chain  $f_1(d) \sqsubseteq f_2(d) \sqsubseteq \dots$ of elements of $\llbracket B \rrbracket$, $\bigsqcup f_n(d)$ is related to that expression. Therefore we have $\bigsqcup f_n(d) \ R_B \ e \ e'$.
%for a chain of elements in $\llbracket A \rrbracket$, $ \bigsqcup x_n R_A e$ and also the functions are continuous,so $f (\bigsqcup x_n) = \bigsqcup (f x_n)$. Therefore we have a chain of elements in $B$ which are the upperBy induction, every element of a chain in $B$ is in
\end{proof}

\section{Main Lemma}
Finally, we can prove the Main Lemma (Lemma \ref{main} above):

\input{mainlem.tex}


%main lemma proof